---
title: "Bayesian implementation"
date: "17/02/2022"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
## Libraries
require(dplyr)
require(data.table)
require(cmdstanr)
melanoma <- boot::melanoma
```

## Maximum likelihood estimation: the Bayesian way

Similar to the frequentist implementation, we use the `melanoma` dataset from the boot package. We load the data and define the competing events.

We use **`CmdStanR`** as the R Stan interface. To install, please follow instructions: <https://mc-stan.org/cmdstanr/>

```{r}
mel_df <- boot::melanoma %>%
  # competing risks
  mutate(comp_event = case_when(
    status %in% c(3) ~ 0, # censored
    status %in% c(2) ~ 2, # progression
    TRUE ~ 1)) %>% # death
  # get the time in year
  mutate(event_time = as.numeric(time/365))
```

```{r}
# Convert to data.table 
df.tmp = as.data.table(mel_df)

# Normalise the age/thickness -- helps HMC sampler, and makes coefficients easier to interpret  
df.tmp[["age"]] = scale(df.tmp[["age"]])
df.tmp[["thickness"]] = scale(df.tmp[["thickness"]])

# Mixture proportions will depend on sex, age, ulcer, thickness 
prop.pars = c("sex", "age", "ulcer", "thickness")

# Construct the matrix of covariates for proportions, adding intercept 
Xprop = cbind(1, as.matrix(df.tmp[,.SD, .SDcols = prop.pars]))
```

```{r}
# Parameters for the risk
pars.2 = c("thickness", "ulcer", "age", "sex")

# Matrices of covariates 
X1 = as.matrix(df.tmp[,.SD, .SDcols = pars.2])
X2 = as.matrix(df.tmp[,.SD, .SDcols = pars.2])

# Bind the two 
X = cbind(X1, X2)

# Save dimensions of each -- we will be passing one big matrix X to Stan, and 
# it needs to figure out which covariates to use for which risk 
cov_dims = t(matrix(c(1, ncol(X1), ncol(X1)+1, ncol(X1) + ncol(X2)), ncol = 2))
```

### Collate the data

The signature should be the same as in the data section of *weibull_mix_cov.stan*

```{r}
data.stan = list(
  N = nrow(mel_df),
  D = length(unique(df.tmp[comp_event != 0,][["comp_event"]])),
  y = mel_df[["event_time"]],
  event = mel_df[["comp_event"]],
  Xprop = Xprop,
  K = length(prop.pars)+1,
  cov_dims = cov_dims,
  M = ncol(X),
  X = X
)
```

### Compile the model with **`CmdStanR`**

We compile the model, however note that we are not sampling at this stage, but creating a model object

```{r warning=FALSE}
mod.cr = cmdstan_model(stan_file = "Bayes/weibull_mix_cov.stan")
```

Now we can sample. To read more about the parameters, please check <https://mc-stan.org/users/documentation/>

```{r message=FALSE, warning=FALSE}
# Sample 
fit.cr <- mod.cr$sample(
  data = data.stan,
  #seed = 123,
  chains = 4,
  parallel_chains = 4,
  refresh = 25,
  max_treedepth = 15,
  adapt_delta = 0.99,
  iter_warmup = 2000,
  iter_sampling = 1000
)
```

```{r}
# Print a summary 
fit.cr$summary()
```

### Results

We can plot the parameters of interest

```{r}
bayesplot::mcmc_hist(fit.cr$draws(c("alpha", "mu", "beta_pi_raw", "beta"))) 
```

For patients with no events recorded (censored), we can calculate mean posterior probability of each event (makes sense in our framework where we select the event first)

```{r}
pr = matrix(apply(fit.cr$draws("pr"), 3, mean), ncol = 2)[df.tmp[["comp_event"]] == 0,]
```

Julia produces extra amazing plots \<3

```{r}
one.chain = as.data.frame(drop(as.array(fit.cr$draws(c("alpha", "mu", "beta_pi_raw", "beta")))[,1,]))
write.csv(one.chain, "one_chain.csv")
```

```{r}
sessionInfo()
```
